Gradual Soundness: Lessons from Static Python

### 2022-01-27 pj-7.1/paper.tex
pj: carl r2 fixes

%% TODO before submission
%% - choose words: static vs typed, nonstatic vs dynamic vs untyped
%%   trying to avoid nonstatic? staticifying?
%% - perhaps a compatibility story, programmers can work on Python 3.9 until Cinder catches up b/c source code is the same
%% - SP doc https://docs.google.com/document/d/1-xzpBuiDYnpeGrXy61R_JCy7-iJ2v6F5DcoB4ICHEfw/edit#
%% -

@@

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}
\newcommand{\PEP}{PEP~484}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\bcinst}[1]{\code{#1}}
\newcommand{\defeq}{=}
\newcommand{\mfeq}{=}
\newcommand{\langmid}{~\mid~} %% {\mathrel{\mathbf{\Big\vert}}}
\newenvironment{langarray}{\(\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
\newcommand{\ipara}[1]{\emph{#1}\/.}
\newcommand{\nnum}[1]{$N=#1$}
\newcommand{\totalnum}[1]{$#1$ total}

%% TODO 2022-01-25 recount tests
\newcommand{\numbenchmark}{2}
\newcommand{\numSPtest}{265}
\newcommand{\numSPtotal}{802} %% SP commit 6d61575 from Jan 9
\newcommand{\numSPdiff}{537} %% (\numSPtotal - \numSPtest)
\newcommand{\CPUchange}{3.7\%}

@@

Unfortunately, the high run-time cost of sound types has split
the gradual typing community.
Industry teams have developed innovative type systems that accommodate
untyped designs, but are unsound~\cite{bat-ecoop-2014,rch-popl-2012,cvgrl-oopsla-2017}.
These systems intentionally check nothing at run-time when untyped values enter
typed code.
Academic teams have primarily focused on the theory of sound
gradual types, formulating correctness properties and studying ever-more-descriptive
types~(e.g.,~\cite{sgt-jfp-2016,clps-popl-2019,nla-popl-2019,mgt-oopsla-2021})
A few academics have studied the cost of run-time checks
in detail~\cite{tfgnvf-popl-2016,gtnffvf-jfp-2019}
and proposed implementation methods~\cite{kas-pldi-2019,bbst-oopsla-2017},
compiler technology~\cite{vsc-dls-2019,fgsfs-oopsla-2018},
and even weakened semantics~\cite{glfd-pj-2022,vss-popl-2017,gi-scp-2020},
but these efforts have not yet decisively closed the performance gap.
The most promising attempt is the \emph{concrete\/} semantics
for gradual types~\cite{clzv-ecoop-2018,mt-oopsla-2017,wzlov-popl-2010}, but it imposes nonlocal
constraints on untyped code.
In particular, concrete-typed client code is incompatible with
values created by untyped code.

@@

\begin{itemize}
  \item
    \emph{Evidence}.
    We present evidence that \SP{} improves performance with few code changes.
    At Instagram, the web server has become significantly more efficient
    after the application of \SP{} to high-profile modules.
    Because the server code is closed-source, we additionally present data for
    \numbenchmark{} public benchmarks.
    Using only shallow types, \SP{} runs on-par with Python despite the costs of
    enforcing soundness.
    With fine-tuned types, \SP{} consistently out-performs the Python baseline.

@@

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Its syntax for types is based on the definitions in \PEP{}~\cite{pep484}.
Its static semantics for types is inspired by optional type checkers; in
particular, Pyre~\cite{pyre} and mypy~\cite{mypy}.
And its strategy for run-time checks is adapted from Nom~\cite{mt-oopsla-2017,mt-oopsla-2021},
a research language with compelling performance results.
\SP{}'s novelty comes from the following engineering constraints:
\begin{itemize}
  \item
    Performance is the bottom line.
    At the end of the day, \SP{} needs to make code run faster in production.
  \item
    Critical code often relies on first-order functions and objects,
    at least in the Instagram web server.
    Precise types for higher-order functions, for first-class classes, and even for
    locally-defined functions and classes are thus a low priority; instead, the
    dynamic type is available as a coarse approximation.
  \item
    The codebase in which \SP{} is being applied makes heavy use of
    \PEP{} type annotations and is regularly analyzed by Pyre.
    \SP{} is therefore compatible with the \PEP{} syntax to reduce the
    adoption burden.
    %% depth-first approach? (the alternatives are not very clear ... depth vs breadth?)
  \item
    A module-level granularity is acceptable.
    Once \SP{} is enabled for a module, it compiles all code in that module
    including expressions that have the dynamic type~(\cref{s:type-dynamic}).
    If this behavior is a problem for certain expressions, it is easy enough to move
    them to an untyped Python module.
\end{itemize}
%
In general, the types that \SP{} implements all enable significant optimizations
and can be validated quickly with casts.
Types that do not meet these criteria are deferred to the mature optional
type checkers that already exist for Python.
In fact, \SP{} has no plans to support all \PEP{} types.
The goal is to implement a small language of types efficiently.

@@

Three other unsupported types bear special mention:
first-class class and object types~(\cref{s:inheritance});
first-class function (callable) types; and
recursive types.
Pyre and mypy some support for the first two kinds, but not for recursive types.
%% mypy rec.types = https://github.com/python/mypy/issues/731
%% nothing in pep 484 = https://www.python.org/dev/peps/pep-0484/

@@

Every \SP{} type corresponds to a constant-time cast.
When an unknown value reaches a boundary to typed code at run-time,
\SP{} uses these checks to quickly decide whether or not the value
matches the type (think: \code{isinstance} checks).
Unlike structural gradual type systems, \SP{} does not traverse
incoming values, nor does it create wrappers to enforce behavior.

These dynamic checks try to faithfully enforce a full type, but some are incomplete---especially
for generic types.
We call the incomplete ones \emph{shallow} types.
The \SP{} strategy for sound generics is to package each value with a type
tag; following the \textsf{StrongScript} authors~\cite{rzv-ecoop-2015}, we call
these \emph{concrete} types.

First off, for a basic nominally-typed language (think Java 1.4, before
generics), constant-time checks are easy to implement.
Every type has a name and every value has a tag that corresponds to the
name of its type.
A cast must examine the parents of a class in the worst case,
but these traversals are bounded and typically short.

Parameterized types (generics) are more difficult to enforce.
For example, suppose that a typed function expects inputs that match
the type \code{Dict[int,int]}; that is, dictionaries with integer keys and
integer values.
A constant-time check for dictionaries is not enough to protect the typed
function because it fails to validate the elements inside a dictionary.
The \SP{} solution is to provide two incompatible versions of every parameterized type,
both of which support efficient run-time checks.\footnote{Not all parameterized \SP{} types have both shallow
and concrete versions at present.
In particular, user-defined classes do not have concrete support.
The shallow check for a user-defined class, however, guarantees the types of all
non-generic fields and methods.}
For dictionaries, \SP{} provides the types $\sptrawpydict$ and $\sptrawchkdict$:
\begin{enumerate}
  \item
    The \emph{shallow} type $\sptrawpydict$ gets enforced with a constant-time check
    for dict values that ignores the elements inside a dict.
    Consequently, a parameterized surface type such as $\sptpydict{K}{V}$
    is merely a promise of dict values at run-time; it does not promise anything about
    keys and values.
  \item
    The \emph{concrete} type $\sptchkdict{K}{V}$ guarantees well-typed
    keys and values, but is incompatible with built-in Python dictionaries.
    Programmers must use a type-annotated contructor to make a concrete-typed checked dict.
\end{enumerate}
%
These options let programmers decide whether to edit their code to support concrete types
or leave it as-is with shallow types.
Often, the shallow types are more attractive because concrete types impose
nonlocal changes.
If one type changes to concrete, then several changes may need to happen:
all values that reach this type must be initialized as concrete,
all typed clients of these values must expect concrete values,
all values that reach those clients must be concrete,
and so on and so forth.

The upside of concrete types is that they enforce stronger type constraints.
These constraints can catch bugs and always lead to more-efficient typed code.
Faster performance is not guaranteed, though, because concrete types must
check all writes from untyped contexts.
\Cref{t:shallow-vs-concrete} presents in more detail the worst-case costs of
shallow and concrete dict types.
Shallow types pay a constant-time cost whenever an untyped value enters typed code
and whenever a typed context reads from a shallow value.
Concrete types pay a potentially-high cost for constructors,
a constant-time cost for casts and for writes in untyped code,
and zero cost for dereferences.
If a value travels to untyped code, gets written to many times, and gets read few times,
then the shallow types may run faster overall.

@@

Cinder adds virtual method tables (vtables) to typed classes.
These tables help to speed up method dispatch relative to Python's
dynamic lookup.
Calls to static methods that appear in statically-typed
code use the vtable to find an address for the method.
If the method is part of a final class, then the call is further optimized
to a direct function-call jump.
(Both vtable lookup and direct jumps are supported by Cinder-specific bytecode
instructions.)
Just-in-time compilation~(\cref{s:jit}) can also upgrade vtable lookups to
function calls.

@@

Python allows classes to inherit from a list of parents.
\SP{} accommodates this behavior provided that one parent is defined in typed
code and the untyped parents do not specify their members (via
\code{\_\_slots\_\_}).
Other uses of multiple inheritance must stay in untyped code.

\SP{} does not leverage types for classes with multiple parents
because of vtable layout issues.
Efficient vtable dispatch requires that each method name
resides in a fixed slot across all subclasses.
With multiple inheritance, however, parents may disagree about which method
lives in each slot.
%% possible improvement: invoke methods on the first parent statically and on
%% other parents dynamically and forbid those other paarents from having instance
%% attributes.

@@

\begin{table}[t]
  \caption{Microbenchmark performance ratios. NB: data for deltablue and nbody is pending.}
  \label{t:microbenchmark}

@@

\SP{} is one of many type systems for Python.
Other \emph{optional\/} type checkers include Pyre~\cite{pyre},
mypy~\cite{mypy}, PyType~\cite{pytype}, and Pyright~\cite{pyright}; these
systems do not ensure safe interactions with untyped Python.
Another sound type system is Reticulated,
which enforces structural types (rather than nominal) using its transient
semantics~\cite{vss-popl-2017,v-thesis-2019}.
Other \emph{optimizing} type systems for Python
include Reticulated and mypyc.
The latter compiles Python source to C extension modules; unlike \SP{},
it does not offer a pure-Python developer experience.

>>>>

%% TODO before submission
%% - choose words: static vs typed, nonstatic vs dynamic vs untyped
%%   trying to avoid nonstatic? staticifying?
%% - perhaps a compatibility story, programmers can work on Python 3.9 until Cinder catches up b/c source code is the same
%% - SP doc https://docs.google.com/document/d/1-xzpBuiDYnpeGrXy61R_JCy7-iJ2v6F5DcoB4ICHEfw/edit#
%% - threats to validity: only 1 migration path, no systematic evaluation (lattice)

@@

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}
\newcommand{\PEP}{PEP~484}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\bcinst}[1]{\code{#1}}
\newcommand{\defeq}{=}
\newcommand{\mfeq}{=}
\newcommand{\langmid}{~\mid~} %% {\mathrel{\mathbf{\Big\vert}}}
\newenvironment{langarray}{\(\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
\newcommand{\ipara}[1]{\emph{#1}\/.}
\newcommand{\nnum}[1]{$N=#1$}
\newcommand{\totalnum}[1]{$#1$ total}
\newcommand{\pctnum}[1]{$#1\%$}

%% TODO 2022-01-25 recount tests
\newcommand{\numbenchmark}{3}
\newcommand{\numSPtest}{265}
\newcommand{\numSPtotal}{802} %% SP commit 6d61575 from Jan 9
\newcommand{\numSPdiff}{537} %% (\numSPtotal - \numSPtest)
\newcommand{\CPUchange}{3.7\%}

@@

Unfortunately, the high run-time cost of sound types has split
the gradual typing community.
Industry teams have developed innovative \emph{optional} type systems that accommodate
untyped designs, but are unsound~\cite{bat-ecoop-2014,rch-popl-2012,cvgrl-oopsla-2017,pep484}.
These systems intentionally check nothing at run-time when untyped values enter
typed code.
Academic teams have primarily focused on the theory of sound
gradual types, formulating correctness properties and studying ever-more-descriptive
types~(e.g.,~\cite{sgt-jfp-2016,clps-popl-2019,nla-popl-2019,mgt-oopsla-2021})
A few academics have studied the cost of run-time checks
in detail~\cite{tfgnvf-popl-2016,gtnffvf-jfp-2019}
and proposed implementation methods~\cite{kas-pldi-2019,bbst-oopsla-2017},
compiler technology~\cite{vsc-dls-2019,fgsfs-oopsla-2018},
and even weakened semantics~\cite{glfd-pj-2022,vss-popl-2017,gi-scp-2020},
but these efforts have not yet decisively closed the performance gap.
The most promising attempt is the \emph{concrete\/} semantics
for gradual types~\cite{clzv-ecoop-2018,mt-oopsla-2017,wzlov-popl-2010}, but it imposes nonlocal
constraints on untyped code.
In particular, concrete-typed client code is incompatible with
values created by untyped code.

@@

\begin{itemize}
  \item
    \emph{Evidence}.
    We present evidence that \SP{} improves performance with few code changes.
    At Instagram, the web server has become significantly more efficient
    after the application of \SP{} to high-profile modules.
    Because the server code is closed-source, we additionally present data for
    \numbenchmark{} public microbenchmarks~(\cref{s:microbenchmarks}).
    Using only shallow types, \SP{} runs slightly faster than Python
    despite the costs of enforcing soundness~(avg. \pctnum{28}).
    With fine-tuned types, \SP{} consistently out-performs the Python baseline~(avg. \pctnum{67}).

@@

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Its syntax for types is based on the definitions in \PEP{}~\cite{pep484}.
Its static semantics for types is inspired by optional type checkers; in
particular, Pyre~\cite{pyre} and mypy~\cite{mypy}.
And its strategy for run-time checks is adapted from Nom~\cite{mt-oopsla-2017,mt-oopsla-2021},
a research language with compelling performance results.
\SP{}'s novelty comes from the following engineering constraints:
\begin{itemize}
  \item
    Performance is the bottom line.
    At the end of the day, \SP{} needs to make code run faster in production.
  \item
    Critical code often relies on first-order functions and objects,
    at least in the Instagram web server.
    Precise types for higher-order functions, for first-class classes, and even for
    locally-defined functions and classes are thus a low priority; instead, the
    dynamic type is available as a coarse approximation.
  \item
    The codebase in which \SP{} is being applied makes heavy use of
    \PEP{} type annotations and is regularly analyzed by Pyre.
    \SP{} is therefore compatible with the \PEP{} syntax to reduce the
    adoption burden.
    %% depth-first approach? (the alternatives are not very clear ... depth vs breadth?)
  \item
    A module-level granularity is acceptable.
    Once \SP{} is enabled for a module, it compiles all code in that module
    including expressions that have the dynamic type~(\cref{s:type-dynamic}).
    If this behavior is a problem for certain expressions, it is easy enough to move
    them to an untyped Python module.
\end{itemize}
%
In general, the types that \SP{} implements all enable significant optimizations
and can be validated quickly with casts.
Types that do not meet these criteria are deferred to the mature optional
type checkers that already exist for Python.
In fact, \SP{} has no short-term plans to support all \PEP{} types.
The main priority is to implement a small language of types efficiently.

@@

Three other unsupported types bear special mention:
first-class class and object types~(\cref{s:inheritance});
first-class function (callable) types; and
recursive types.
Pyre and mypy offer some support for the first two kinds, but not for recursive types.
%% mypy rec.types = https://github.com/python/mypy/issues/731
%% nothing in pep 484 = https://www.python.org/dev/peps/pep-0484/

@@

Every \SP{} type is paired with an efficient cast operation that runs in
constant time regardless of how large an incoming value is.
Unlike structural gradual type systems~\cite{gtnffvf-jfp-2019},
no cast traverses an incoming value and no cast creates a wrapper
to monitor future behaviors.
Some casts for generic types are, however, incomplete.
We call the types with incomplete casts \emph{shallow} types
and the types with full casts \emph{concrete} types, following
the \code{StrongScript} authors~\cite{rzv-ecoop-2015}.

First off, the casts for non-generic nominal types are straightforward.
Every type has a name and every value has a tag that corresponds to the name.
A cast must examine the parents of a class to find a match in the worst case,
but these traversals are bounded in length and typically short.

Generic types are more difficult to enforce because a tag by itself does not
describe their contents.
For example, suppose that a function expects inputs that match
the type \code{Dict[int,int]}; that is, dictionaries with integer keys and
integer values.
A constant-time check for Python dictionaries is not enough to protect the
function against dicts with ill-typed elements.
The \SP{} solution is provide a second kind of dictionary that ensures
the validity of its elements.
These two dicts have distinct types:
\begin{enumerate}
  \item
    The \emph{shallow} type $\sptrawpydict$ gets enforced with a constant-time check
    for dict values that ignores elements.
    Consequently, a parameterized surface type such as $\sptpydict{K}{V}$
    does not promise anything about keys and values.
  \item
    The \emph{concrete} type $\sptchkdict{K}{V}$ guarantees well-typed
    keys and values, but is incompatible with built-in Python dictionaries.
    Programmers must use a type-annotated contructor to make a concrete-typed checked dict.
\end{enumerate}
%
These options let programmers decide whether to edit their code to support concrete types
or leave it as-is with shallow types.\footnote{Not all generic \SP{} types have shallow
  and concrete versions at present.
  In particular, user-defined classes do not have concrete support.
  The shallow check for a user-defined class does, however, guarantee the types of all
  non-generic fields and methods.}
Often, the shallow types are more attractive because concrete types impose
nonlocal changes.
If one type changes to concrete, then several changes may need to happen:
all values that reach this type must be initialized as concrete,
all typed clients of these values must expect concrete values,
all values that reach those clients must be concrete,
and so on and so forth.

The upside of concrete types is that they enforce stronger type constraints.
These constraints can catch bugs and always lead to faster typed code.
Faster performance is not guaranteed in general, though, because concrete types must
check all writes from untyped contexts.
\Cref{t:shallow-vs-concrete} presents in more detail the worst-case costs of
shallow and concrete dict types.
Shallow types pay a constant-time cost whenever an untyped value enters typed code
and whenever a typed context reads from a shallow value.
Concrete types pay a potentially-high cost for constructors,
a constant-time cost for casts and for writes in untyped code,
and zero cost for reads.
If a value travels to untyped code, gets written to many times, and gets read few times,
then the shallow types may run faster in the end.

@@

%% vtables go in:
%% - typed classes
%% - builtin classes that SP treats as typed
%% - typed and untyped subtypes of classes with vtables

Cinder adds virtual method tables (vtables) to classes.
These tables help to speed up method dispatch relative to Python's
dynamic lookup.
Calls to static methods that appear in statically-typed
code use the vtable to find an address for the method.
If the method is part of a final class, then the call is further optimized
to a direct function-call jump.
(Both vtable lookup and direct jumps are supported by Cinder-specific bytecode
instructions.)
Just-in-time compilation~(\cref{s:jit}) can also upgrade vtable lookups to
function calls.

@@

Python allows classes to inherit from a list of parents.
\SP{} accommodates this behavior provided that at most one parent
is typed (or uses \code{\_\_slots\_\_} to specify its layout; plain Python has the same restriction).

\SP{} does not leverage types for classes with multiple parents
because of vtable layout issues.
Efficient vtable dispatch requires that each method name
resides in a fixed slot across all subclasses.
With multiple inheritance, however, parents may disagree about which method
lives in each slot.
Similarly, \SP{} cannot optimize instance variable reads because their slots
may have a conflict among parents.
%% possible improvement: invoke methods on the first parent statically and on
%% other parents dynamically and forbid those other paarents from having instance
%% attributes.

@@

\begin{table}[t]
  \caption{Microbenchmark performance ratios. NB: data for nbody is pending.}
  %% FILL don't forget to update \numbenchmark{}
  \label{t:microbenchmark}

@@

\SP{} is one of many type systems for Python.
Other optional type checkers include Pyre~\cite{pyre},
mypy~\cite{mypy}, PyType~\cite{pytype}, and Pyright~\cite{pyright}; these
systems do not ensure safe interactions with untyped Python.
Another sound type system is Reticulated,
which enforces structural types (rather than nominal) using its transient
semantics~\cite{vss-popl-2017,v-thesis-2019}.
Other \emph{optimizing} type systems for Python
include Reticulated and mypyc.
The latter compiles Python source to C extension modules; unlike \SP{},
it does not offer a pure-Python developer experience.
